---
title: "Analyzing qPCR Data: Better Practices to Facilitate Rigor and Reproducibility - Figures and Tables Code"
output: html_notebook
---


```{r message=F, warning=F}
library(readxl)
library(tidyverse)
library(qpcR)
library(rentrez)
library(gtools)
```

# Load Data

Load data used to create the figures and run the models. 

```{r}
qpcr_results_files <- c("./data/2024-03-07_GAPDH.xls","./data/2023-07-31_MMP10.xls")
qpcr_metadata_files <- c("./data/Experiment Key GAPDH.xlsx","./data/Experiment Key MMP10.xlsx")
files_list <- list("GAPDH" = c(qpcr_results_files[1], qpcr_metadata_files[1]),
                   "MMP10" = c(qpcr_results_files[2], qpcr_metadata_files[2]))
```

Our lab uses the qPCR QuantStudio 3 machine running the QuantStudio Design and Analysis software v1.5.2, which generates results in Excel Workbooks. Each workbook has an "Amplification Data" sheet which contains the amplification information for each well on a 96-well plate (see `2024-03-07_GAPDH.xls` and `2023-07-31_MMP10.xls`). 

The `ReadExcelAmplification` function extracts the amplification data from the generated Excel Workbooks and joins it with experimental metadata. In this worked example, metadata is provided in the `Experiment Key GAPDPH.xlsx` and `Experiment Key MMP10.xlsx` files.

```{r}
ReadExcelAmplifcation <- function(results_file, metadata_file){
  # read data from provided excel spreadsheets
  amp.curves <- read_excel(results_file, sheet = "Amplification Data", skip = 42, .name_repair = make.names)
  metadata <- read_excel(metadata_file, skip = 10)
  # join amplification data with metadata
  amp.curves.annotated <- amp.curves |> 
    inner_join(metadata, by=join_by("Well.Position"=="QPCR Well"))
  
  return(amp.curves.annotated)
}
```


```{r}
# read in amplification data
amp.curves.annotated <- lapply(files_list, function(x){
  ReadExcelAmplifcation(x[1], x[2])
})
names(amp.curves.annotated) <- names(files_list)
```

# Exploring Fluorescence Curves

## Figure Theme
```{r}
figure_theme <- theme(axis.text = element_text(size = 18),
                      axis.title = element_text(size = 20),
                      legend.text = element_text(size = 18),
                      legend.title = element_text(size = 20),
                      panel.background = element_rect(fill = "white", colour = NA), 
                      panel.border = element_rect(fill = NA, colour = "grey20"), 
                      panel.grid = element_line(colour = "grey92"), 
                      panel.grid.minor = element_line(linewidth = rel(0.5)), 
                      strip.background = element_rect(fill = "grey85", colour = "grey20"), 
                      complete = TRUE)
```

Create plotting dataframe.

```{r}
# GAPDH
amp.curves.annotated$GAPDH$group <- paste(amp.curves.annotated$GAPDH$Donor, amp.curves.annotated$GAPDH$Treatment, sep = '_')

GAPDH_mean_delta_df <- amp.curves.annotated$GAPDH %>%
  group_by(Cycle, group) %>%  # Group by the 'group' column
  summarise(Mean_Delta = mean(Delta.Rn))   
 
GAPDH_mean_delta_df$Donor <- unlist(lapply(GAPDH_mean_delta_df$group, function (g){
  unlist(strsplit(g, '_'))[1]
}))

GAPDH_mean_delta_df$Treatment <- unlist(lapply(GAPDH_mean_delta_df$group, function (g){
  unlist(strsplit(g, '\\+ '))[2]
}))

# MMP10

amp.curves.annotated$MMP10$group <- paste(amp.curves.annotated$MMP10$Donor, amp.curves.annotated$MMP10$Treatment, sep = '_')

MMP10_mean_delta_df <- amp.curves.annotated$MMP10 %>%
  group_by(Cycle, group) %>%  # Group by the 'group' column
  summarise(Mean_Delta = mean(Delta.Rn))   
 
MMP10_mean_delta_df$Donor <- unlist(lapply(MMP10_mean_delta_df$group, function (g){
  unlist(strsplit(g, '_'))[1]
}))

MMP10_mean_delta_df$Treatment <- unlist(lapply(MMP10_mean_delta_df$group, function (g){
  unlist(strsplit(g, '\\+ '))[2]
}))
```

## Figure 1A

```{r}
Trips <- amp.curves.annotated$GAPDH[amp.curves.annotated$GAPDH$Donor %in% c("KK32G","KK29H") &
                             amp.curves.annotated$GAPDH$Treatment ==  "ASM Media Control + DMSO", ]
ggplot(Trips, aes(x = Cycle, y = Delta.Rn, colour = Well.Position,
       linetype = Donor)) +
  geom_line() +  # Use geom_line for a line plot
  labs(
    x = "Cycle",
    y = "GAPDH Fluorescence",
    title = ""
  ) +
  theme_set(figure_theme)

ggsave("./Figures/GAPDH_Control_2Donors.pdf", width = 7, height = 5, units = "in")
```

## Figure 1B

Plot All GAPDH Average Curves 

```{r}
ggplot(GAPDH_mean_delta_df, aes(x = Cycle, y = Mean_Delta, colour = Treatment, 
       linetype = Donor)) +
  geom_line() +  # Use geom_line for a line plot
  labs(
    x = "Cycle",
    y = "GAPDH Fluorescence",
    title = ""
  ) + 
  xlim(19,39) +
  theme_set(figure_theme)
ggsave("./Figures/GAPDH_average_cuves.pdf", width = 7, height = 5, units = "in")
```

## Figure 1C

Plot average MMP10 curves for all donors
```{r}
ggplot(MMP10_mean_delta_df, aes(x = Cycle, y = Mean_Delta, colour = Treatment, 
       linetype = Donor)) +
  geom_line() +  # Use geom_line for a line plot
  labs(
    x = "Cycle",
    y = "MMP10 Fluorescence",
    title = ""
  ) + xlim(19,39) +
  theme_set(figure_theme) +
  geom_hline(yintercept = 2, linetype = "dashed")  # Add the horizontal line

ggsave("./Figures/MMP10_average_cuves.pdf", width = 7, height = 5, units = "in")
```

## Figure 1D

Predict Fluoresence Values for Mean MMP10 Curves

Curve fit function.

```{r}
# This function assumes fluorescence data are in columns and rows 
# are cycles. 
FitCurve <- function(amp.data){
  # run pcrfit
  pcrfit.models <- lapply(2:ncol(amp.data), function (well){
    pcrfit(as.data.frame(amp.data), 1, well, model = l4)
  })
  well_ids <- unlist(lapply(pcrfit.models, function(m){m[["names"]]}))
  names(pcrfit.models) <- well_ids
  
  return(pcrfit.models)
}
```

Mean fluorescence in each group

```{r}
# make MMP10_mean_delta_df wide
MMP10_mean_f <- amp.curves.annotated$MMP10 %>%
  group_by(group, Cycle) %>%
  summarize(mean_Delta.Rn = mean(Delta.Rn))


MMP10_mean_f.wide <- MMP10_mean_f |> 
  pivot_wider(id_cols=Cycle,
              names_from = group,
              values_from = mean_Delta.Rn)
```

Fit mean fluorescence at each cycle by group

```{r}
MMP10_mean_fits <- FitCurve(MMP10_mean_f.wide)

MMP10_mean.Donor <- unlist(lapply(names(MMP10_mean_fits), function (n){
  unlist(strsplit(n, '_'))[1]
}))

MMP10_mean.Treatment <- unlist(lapply(names(MMP10_mean_fits), function (n){
   unlist(strsplit(n, '\\+ '))[2]
}))

MMP10_mean.Treatment <- sub("_+$", "", MMP10_mean.Treatment)
```

Predict point at which fluorescence reaches 2

```{r}
myCycles <- seq(from = 27, to = 38, by = 0.1)

MMP10pred2 <- lapply(MMP10_mean_fits, function (m){
  P <- predict(m, newdata = data.frame(Cycles = myCycles))
  myCycles[which(P>2)[1]]
})

MMP10_Results <- data.frame("T" = unlist(MMP10pred2),
                            "Donor"  = MMP10_mean.Donor,
                            "Treatment" = MMP10_mean.Treatment
                            )
```

Plot

```{r}
ggplot(MMP10_Results, aes(x = Treatment, y = T, shape = Donor, group = Donor)) +  
  geom_line() +  # Create line plots
  geom_point() + # Add points to make it easier to see the  data
  labs(
    x = "", 
    y = "Cycle", 
    title = ""
  ) +
  ylim(29, 37) +
  theme_set(figure_theme) +
  geom_point(size=4) 
ggsave("./Figures/MMP10_Rn2.pdf", width = 7, height = 5, units = "in")
```

Simple statistics: Paired t test

```{r}
PairedMMP10 <- t.test(MMP10_Results$T[MMP10_Results$Treatment == "ETI"],
       MMP10_Results$T[MMP10_Results$Treatment == "DMSO"],
       paired = TRUE)

PairedMMP10
```

Unpaired t test

```{r}
unPairedMMP10 <- t.test(MMP10_Results$T[MMP10_Results$Treatment == "ETI"],
       MMP10_Results$T[MMP10_Results$Treatment == "DMSO"],
       paired = FALSE)

unPairedMMP10
```

# Manualy Estimating Efficiency from Fluorescence Slopes

## Figure 2A

```{r}
MMP10_meanExpo <- MMP10_mean_delta_df[MMP10_mean_delta_df$Cycle > 23 & 
                                      MMP10_mean_delta_df$Cycle < 29 , ]

MMP10_meanExpo$log2F <- log2(MMP10_meanExpo$Mean_Delta)

ggplot(MMP10_meanExpo, aes(x = Cycle, y = log2F, 
                           colour = Treatment, 
       shape = Donor)) +
  geom_point() +  # Use geom_line for a line plot
  labs(
    x = "Cycle",
    y = "Delta Rn (log2)"
  ) +
  geom_point(size=4) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_set(figure_theme) 

ggsave("./Figures/MMP10_Exponential.pdf", width = 7, height = 5, units="in")
```

## Figure 2B

Slope Differences


```{r}
MMP10_eff <- sapply(unique(MMP10_meanExpo$group), function (g){
  r <- coefficients(summary(lm( log2F ~ Cycle, data = MMP10_meanExpo[MMP10_meanExpo$group == g,])))
  r["Cycle", "Estimate"]
})

names(MMP10_eff)<- str_replace(unique(MMP10_meanExpo$group), "_.*\\+ ", "_") 

# Split the strings on the underscore
split_strings <- str_split(names(MMP10_eff), "_")

MMP10_eff_df <- data.frame("Efficiency" = MMP10_eff,
"Donor" =  sapply(split_strings, function(x) x[1]),
"Treatment" = sapply(split_strings, function(x) x[2])
)

ggplot(MMP10_eff_df, aes(x = Donor, y = Efficiency, fill = Treatment)) +  # Fill by 'Treatment'
  geom_bar(stat = "identity", position = "dodge") +  # Use 'dodge' for side-by-side bars
  labs(
    x = "Donor", 
    y = "Slope", 
  ) +
  theme(figure_theme)+
  theme(axis.text.x = element_text(angle = 45))
 
ggsave("./Figures/EffEstSlope.pdf", width = 7, height = 5, units = "in")
```

# PubMed Central References to qPCR


```{r}
# Function to search PubMed Central and count "qPCR" mentions in methods sections
search_pmc <- function(year) {
  # Search term for methods section mentioning qPCR in a specific year
  search_term <- paste0("(\"methods\"[Section] AND qPCR) AND ", year, "[PDAT]")

  # Search PubMed Central
  search_result <- entrez_search(db = "pmc", term = search_term)

  # Return the count of articles found
  return(search_result$count)
}

# Years to search (e.g., 2010 to 2023)
years <- 2010:2023

# Create an empty data frame to store results
results <- data.frame(Year = integer(), Count = integer())

# Loop through each year and perform the search
for (year in years) {
  count <- search_pmc(year)
  results <- rbind(results, data.frame(Year = year, Count = count))
  Sys.sleep(1)  # Pause to avoid overloading the server
}

# Print the results
print(results)
```

## Figure 3A

```{r}
# Optional: Plot the results using ggplot2
ggplot(results, aes(x = Year, y = Count)) +
  geom_col() +
  labs(x = "", 
       y = "Publications") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2010, 2023, 3))+
  theme(axis.text.x = element_text(angle = 45))

ggsave("./Figures/qPCR.Ref.pdf", width = 7, height = 5, units = "in")
```

## Figure 3B

```{r}
DDCTvPfaffl <- read_excel("./data/DDCTvPfaffl.xlsx")
# Reshape the data for ggplot
df_long <- DDCTvPfaffl %>% 
  pivot_longer(cols = c("DDCT", "Pfaffl"), names_to = "Method", values_to = "Value")

# Create the ggplot
ggplot(df_long, aes(x = Year, y = Value, fill = Method)) +
  geom_col(position = "dodge") +  # Use 'dodge' for side-by-side bars
  labs(
    x = "", 
    y = "Citations") +
  theme_set(figure_theme) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_discrete(
    labels = c(expression(paste(Delta,Delta, "Ct")), "Pfaffl") 
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))   

ggsave("./Figures/DDCTvPfaffl.pdf", width = 7, height = 5, units = "in")
```

# Amplification Efficiency

Plots showing amplification efficiency for each well on a qPCR plate or for each replicate group.

```{r}
# format data for GAPDH and MMP10
GAPDH.amp.curves.annotated.wide <- amp.curves.annotated$GAPDH |> 
  pivot_wider(id_cols = Cycle,
              names_from = Well.Position,
              values_from = Delta.Rn)
MMP10.amp.curves.annotated.wide <- amp.curves.annotated$MMP10 |> 
  pivot_wider(id_cols=Cycle,
              names_from = Well.Position,
              values_from = Delta.Rn)

# format data with replicates
amp.curves.annotated.replicates <- lapply(amp.curves.annotated, function(gene.data){
  # get replicate IDs from metadata
  replicate.ids <- unique(gene.data$`Sample Number for qPCR`)
  # generate wide-format data frame from each replicate group
  replicate.list <- lapply(replicate.ids, function(rep){
    rep.data <- gene.data |> 
      filter(`Sample Number for qPCR` == rep) |> 
      pivot_wider(id_cols = Cycle,
                  names_from = Well.Position,
                  values_from = Delta.Rn)
    return(rep.data)
  })
  names(replicate.list) <- replicate.ids
  return(replicate.list)
})
names(amp.curves.annotated.replicates) <- names(amp.curves.annotated)
```

Helper functions to get efficiency data and create graphs.

```{r}
GetEfficiencyEstimates <- function(amp.data){
  # run pcrfit
  pcrfit.models <- lapply(2:ncol(amp.data), function (well){
    pcrfit(as.data.frame(amp.data), 1, well, model = l4)
  })
  well_ids <- unlist(lapply(pcrfit.models, function(m){m[["names"]]}))
  names(pcrfit.models) <- well_ids
  
  # extract efficiency estimates
  eff.estimates <- sapply(pcrfit.models, function (m){
    efficiency(m, method = "spline", plot = FALSE)[["eff"]]
  }, USE.NAMES = T, simplify = "array")
  
  return(eff.estimates)
}

GetEfficiencyEstimates.Replicates <- function(amp.data.replicates){
  # run pcrfit for each batch of replicates
  pcrfit.models <- lapply(amp.data.replicates, function (batch){
    pcrfit(as.data.frame(batch), 1, 2:ncol(batch), model = l4)
  })
  names(pcrfit.models) <- names(amp.data.replicates)
  
  # extract efficiency estimates
  eff.estimates <- sapply(pcrfit.models, function (m){
    efficiency(m, method = "spline", plot = FALSE)[["eff"]]
  }, USE.NAMES = T, simplify = "array")
  
  return(eff.estimates)
}

GraphEfficiencyEstimatesColored <- function(plot.df, ylab="Well ID"){
  
  stats.df <- data.frame(Mean=mean(plot.df$Eff), 
                         Lower = mean(plot.df$Eff)-sd(plot.df$Eff),
                         Upper = mean(plot.df$Eff)+sd(plot.df$Eff),
                         ypos = levels(plot.df$Well)[floor(nrow(plot.df)/2)])
  
  p.obj <- ggplot(data=plot.df, aes(y=Well, x=Eff, color=Donor, shape=Treatment))+
    geom_point(size=2.5)+
    geom_vline(xintercept = c(1,2), color="black", linetype="dashed")+
    geom_point(data=stats.df, aes(x=Mean, y=ypos), color="red",size=2, alpha =0.8, inherit.aes = F, show.legend = F)+
    geom_errorbar(data=stats.df, aes(y=ypos, xmax = Upper, xmin = Lower), inherit.aes = F, color="red", show.legend = F)+
    annotate("text",
             y=stats.df$ypos, 
             x=stats.df$Upper+0.2, 
             label=paste0("Mean(eff):",round(stats.df$Mean,2), "\n", "Std. Dev.(eff):", round(stats.df$Upper - stats.df$Mean,2)),
             size = 4)+
    theme_minimal()+
    theme(legend.position = "right", 
          legend.justification.right = "top",
          plot.background = element_rect(fill="white"),
          axis.text = element_text(size=12, color = "black"),
          legend.text = element_text(size=18),
          legend.title = element_text(size=20))+
    labs(x="Efficiency Estimate",
         y=ylab,
         color = "Donor")
  return(p.obj)
}
```

## Figure 4A

GAPDH individual well amplification efficiencies. 

```{r}
GAPDH.eff.estimates <- GetEfficiencyEstimates(GAPDH.amp.curves.annotated.wide)
figA.df <- data.frame(Well=factor(names(GAPDH.eff.estimates), levels=rev(mixedsort(names(GAPDH.eff.estimates)))), 
                      Eff = GAPDH.eff.estimates)

# Join in metadata
GAPDH.metadata <- read_excel(qpcr_metadata_files[1], skip=10)
GAPDH.metadata <- GAPDH.metadata |> 
  dplyr::select(Donor, Treatment, `QPCR Well`) |> 
  rename(WellID = `QPCR Well`) |> 
  mutate(Treatment = str_split_i(Treatment, " \\+ ", 2)) |> 
  mutate(Label = paste(Donor, Treatment, sep = " "))

figA.plot <- figA.df |> 
  left_join(GAPDH.metadata, by=join_by("Well"=="WellID")) |> 
  mutate(Well=factor(Well, levels = rev(mixedsort(Well))))

# Create graph
GraphEfficiencyEstimatesColored(figA.plot)+
  labs(title="GAPDH Efficiency")
ggsave("./Figures/Figure4A.pdf", width = 7, height = 6, units = "in")
```

## Figure 4B

MMP10 individual well amplification efficiencies. 

```{r}
MMP10.eff.estimates <- GetEfficiencyEstimates(MMP10.amp.curves.annotated.wide)
figB.df <- data.frame(Well=factor(names(MMP10.eff.estimates), levels=rev(mixedsort(names(GAPDH.eff.estimates)))), 
                      Eff = MMP10.eff.estimates)

# Join in metadata
MMP10.metadata <- read_excel(qpcr_metadata_files[2], skip=10)
MMP10.metadata <- MMP10.metadata |> 
  dplyr::select(Donor, Treatment, `QPCR Well`) |> 
  rename(WellID = `QPCR Well`) |> 
  mutate(Treatment = str_split_i(Treatment, " \\+ ", 2)) |> 
  mutate(Label = paste(Donor, Treatment, sep = " "))

figB.plot <- figB.df |> 
  left_join(MMP10.metadata, by=join_by("Well"=="WellID")) |> 
  mutate(Well=factor(Well, levels = rev(mixedsort(Well))))

# Create graph
GraphEfficiencyEstimatesColored(figB.plot)+
  labs(title="MMP10 Efficiency")
ggsave("./Figures/Figure4B.pdf", width = 7, height = 6, units = "in")
```

## Figure 4C

GAPDH replicate group amplification efficiencies. 

```{r}
GAPDH.eff.estimates.reps <- GetEfficiencyEstimates.Replicates(amp.curves.annotated.replicates$GAPDH)

figC.df <- data.frame(Well=factor(names(GAPDH.eff.estimates.reps), levels=rev(mixedsort(names(GAPDH.eff.estimates.reps)))), 
                      Eff = GAPDH.eff.estimates.reps)

# Join in metadata
partC.metadata <- amp.curves.annotated[["GAPDH"]] |> 
  dplyr::select(`Sample Number for qPCR`, Treatment, Donor) |> 
  distinct() |> 
  mutate(Treatment = str_split_i(Treatment, " \\+ ", 2)) |> 
  mutate(Label = paste(Donor, Treatment, sep=" ")) |> 
  mutate(Well  = factor(`Sample Number for qPCR`))

figC.plot <- figC.df |> 
  left_join(partC.metadata, by=join_by(Well))

# Create graph
GraphEfficiencyEstimatesColored(figC.plot, ylab="Replicate Group")+
  labs(title = "GAPDH Efficiency")
ggsave("./Figures/Figure4C.pdf", width = 7, height = 5, units = "in")
```

## Figure 4D

MMP10 replicate group amplification efficiencies. 

```{r}
MMP10.eff.estimates.rep <- GetEfficiencyEstimates.Replicates(amp.curves.annotated.replicates$MMP10)
figD.df <- data.frame(Well=factor(names(MMP10.eff.estimates.rep), levels=rev(mixedsort(names(MMP10.eff.estimates.rep)))), 
                      Eff = MMP10.eff.estimates.rep)

# Join in metadata
partD.metadata <- amp.curves.annotated[["MMP10"]] |> 
  dplyr::select(`Sample Number for qPCR`, Treatment, Donor) |> 
  distinct() |> 
  mutate(Treatment = str_split_i(Treatment, " \\+ ", 2)) |> 
  mutate(Label = paste(Donor, Treatment, sep=" ")) |> 
  mutate(Well  = factor(`Sample Number for qPCR`))

figD.plot <- figD.df |> 
  left_join(partD.metadata, by=join_by(Well))

# Create graph
GraphEfficiencyEstimatesColored(figD.plot, ylab="Replicate Group")+
  labs(title = "MMP10 Efficiency")
ggsave("./Figures/Figure4D.pdf", width = 7, height = 5, units = "in")
```

# Reference and Target Cycle Thresholds 

Visualize whether treatment effects CT of reference genes.

```{r}
ReadExcelCTResults <- function(results_file, metadata_file){
  results <- read_excel(results_file, sheet = "Results", skip = 42, .name_repair = make.names)
  metadata <- read_excel(metadata_file, skip = 10)
  
  results.merged <- results |> 
    inner_join(metadata, by=join_by("Well.Position" == "QPCR Well"))
  
  return(results.merged)
}
```


```{r}
ct.results.annotated <- lapply(files_list, function(x){
  ReadExcelCTResults(x[1],x[2])
})
```

Helper functions to format CT data for plotting and modeling and to graph CT data.

```{r}
FormatCT.ANCOVA <- function(CTResults, SampleCol="Sample.Name", DonorCol="Donor", TreatmentCol="Treatment", TargetCol="Target.Name", CTCol="Ct.Mean"){
  formatted <- CTResults |> 
    dplyr::select(all_of(c(SampleCol, DonorCol, TreatmentCol, TargetCol, CTCol))) |> 
    distinct()
  return(formatted)
}

JoinCTResults <- function(CT1, CT2, TreatmentCol="Treatment", TargetCol="Target.Name", DonorCol="Donor", CTCol = "Ct.Mean"){
  CT1$Group <- paste0(CT1[[DonorCol]]," ",CT1[[TreatmentCol]])
  CT2$Group <- paste0(CT2[[DonorCol]], " ", CT2[[TreatmentCol]])
  CT1.gene <- CT1[[TargetCol]][1]
  CT2.gene <- CT2[[TargetCol]][1]
  
  joined.results <- CT1[,c(DonorCol, TreatmentCol, "Group", CTCol)] |> 
    inner_join(CT2[,c("Group",CTCol)], by=join_by(Group), suffix = c(paste0(".",CT1.gene), paste0(".",CT2.gene)))
  
  return(joined.results)
}

GraphCT <- function(CTData, CT1Col="Ct.Mean.GAPDH", CT2Col="Ct.Mean.MMP10", TreatmentCol="Treatment", DonorCol="Donor"){
  plot.df <- CTData |> 
    pivot_longer(cols = all_of(c(CT1Col, CT2Col)),
                 values_to = "CT",
                 names_to = "Gene")
  plot.df$LineGroup <- paste0(plot.df[[DonorCol]]," ",plot.df$Gene)
  
  # rename treatment col for group_by access
  plot.df$AccessTreatment <- plot.df[[TreatmentCol]]
  # calculate average CT for each gene across each treatment group
  average.df <- plot.df |> 
    group_by(AccessTreatment, Gene) |> 
    summarise(meanCT = mean(CT)) |> 
    rename(AvgLineGroup=Gene)
  
  # plot CT by treatment group, colored by Donor
  p.obj <- ggplot(plot.df, aes_string(x=TreatmentCol, y="CT"))+
    geom_point(aes_string(shape="Gene", color = DonorCol), size=4)+
    geom_line(aes_string(group = "LineGroup", color = DonorCol))+
    theme_set(figure_theme)
  
  # add lines representing average CT for each gene / treatment combination
  p.obj <- p.obj+
    geom_point(data=average.df, aes(x=AccessTreatment, y=meanCT, shape=AvgLineGroup), size=4, inherit.aes = F)+
    geom_line(data=average.df, aes(x=AccessTreatment, y=meanCT, group=AvgLineGroup), inherit.aes = F, linetype="dashed")
    
  return(p.obj)
}
```

Format CT data. 

```{r}
GAPDH.CT <- FormatCT.ANCOVA(ct.results.annotated$GAPDH)
MMP10.CT <- FormatCT.ANCOVA(ct.results.annotated$MMP10)

# reformat treatment column
GAPDH.CT$Treatment <- gsub(" ","",str_split_i(GAPDH.CT$Treatment,"\\+",2))
MMP10.CT$Treatment <- gsub(" ","",str_split_i(MMP10.CT$Treatment,"\\+",2))

# join data
CT.joined <- JoinCTResults(GAPDH.CT, MMP10.CT)
```


## Figure 5A

CT values for GAPDH and MMP10 stratified by donor and treatment.

```{r}
GraphCT(CT.joined)
ggsave("./Figures/Figure5A.pdf")
```

## Figure 5B

Correlation between GAPDH and MMP10

```{r}
CT.joined.Means <- CT.joined

colnames(CT.joined.Means) <- c("Donor","Treatment","Group", "GAPDH", "MMP10")

Cor.plot.df <-  pivot_longer(CT.joined.Means, cols = c("GAPDH", "MMP10"),
                             values_to = "CT",
                             names_to = "Gene")

ggplot(CT.joined.Means, aes(x=GAPDH, y=MMP10, col=Donor, shape = Treatment) ) +
  geom_point(size = 4) +
  geom_smooth(aes(group=1), method = "lm", formula = y ~ x, se = FALSE, color= "black") +
  theme_set(figure_theme)

ggsave("./Figures/Figure5B.pdf")
```
## Figure 5C

CT values for DEFB1 and HSP90AB stratified by donor and treatment.

```{r}
CT_mat <- read.csv(file="./data/CTmat.csv")
All.CT <- merge(CT.joined.Means[, c("Donor", "Treatment", "Group")],  CT_mat, by.x = "Group", by.y = "X") 
GraphCT(All.CT, CT1Col="HSP90AB", CT2Col="DEFB1", TreatmentCol="Treatment", DonorCol="Donor")
ggsave("./Figures/Figure5C.pdf")
```

## Figure 5D

Correlation between GAPDH and MMP10

```{r}
ggplot(All.CT, aes(x=HSP90AB, y=DEFB1, col=Donor, shape = Treatment) ) +
  geom_point(size = 4) +
  geom_smooth(aes(group=1), method = "lm", formula = y ~ x, se = FALSE, color= "black") +
  theme_set(figure_theme)

ggsave("./Figures/Figure5D.pdf")
```

# ANCOVA Modeling

Correlation between GPDH and MMP10 CT values

```{r}
GAPDH.MMP10.corr <- cor.test(CT.joined.Means$GAPDH, CT.joined.Means$MMP10)

GAPDH.MMP10.rsq <- GAPDH.MMP10.corr$estimate^2
GAPDH.MMP10.corr
```
Once the data are in the right format, ANCOVA is easy. ETI matters.

## Table 2

Data in format ready for modeling.

```{r}
CT.joined.Means
```
## MMP10 Model

Model MMP10 as a function of donor, treatment, and reference gene GAPDH.

```{r}
MMP10.Donor.ETI.GAPDH.fit <- lm(MMP10 ~ Donor + Treatment + GAPDH, data = CT.joined.Means)
summary(MMP10.Donor.ETI.GAPDH.fit)
```

### Table 3

MMP10 modeling results. 

```{r}
MMP10_digest <- summary(MMP10.Donor.ETI.GAPDH.fit)$coefficients

# Create a data frame from the summary
MMP10_digest <- data.frame(
  Term = rownames(MMP10_digest)[-1],
  Estimate = MMP10_digest[-1, "Estimate"],
  p_value = MMP10_digest[-1, "Pr(>|t|)"],
  row.names = NULL  # Remove row names for cleaner output
)
MMP10_digest
```

```{r}
write.csv(MMP10_digest, file = "MMP10_digest.csv", row.names = FALSE)
```


## DEFB1 Model

ANCOVA of DEFB1 using HSP90AB as a reference

```{r}
DEFB1.Donor.ETI.HSP90AB.fit <- lm(DEFB1 ~ Donor + Treatment + HSP90AB, data = All.CT)
summary(DEFB1.Donor.ETI.HSP90AB.fit)
```

### Table 4

```{r}
DEFB1_digest <- summary(DEFB1.Donor.ETI.HSP90AB.fit)$coefficients

# Create a data frame from the summary
DEFB1_digest <- data.frame(
  Term = rownames(DEFB1_digest)[-1],
  Estimate = DEFB1_digest[-1, "Estimate"],
  p_value = DEFB1_digest[-1, "Pr(>|t|)"],
  row.names = NULL  # Remove row names for cleaner output
)

DEFB1_digest
```

```{r}
write.csv(DEFB1_digest, file = "DEFB1_digest.csv", row.names = FALSE)
```